Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-03-31 ms.tex
blah blah

For low-mass black holes, stellar winds are the most important source of heating on
timescales of less than a few tens of millions of years.  Compton
heating comes to dominate at later times due to the higher accretion
rates that accompany the overall decrease in the heating rate, coupled
with the persistently high mass loss rates and gas densities associated with a relatively young
stellar population.  Compton heating dominates at all epochs for
higher mass black holes, with stellar winds and Ia SN competing for
the most significant non-AGN heating source.  As the total heating
rate declines with time, after some point the flow becomes thermally
unstable according to the criterion $(\dot{q}_{\rm heat}/|\dot{q}_{\rm
  rad}|)_{r_{s}} \lesssim 1$ (eq.~[\ref{eq:cooling3}]), as shown by
dashed lines in Fig.~\ref{fig:vwSourcesImp}.  The thermal stability is
calculated neglecting Compton heating because the latter may not be
capable of truly stabilizing the flow given its inability to respond
instantaneously to local changes in the properties of the gas.
Thermally unstable flow is seen to be the norm in the single starburst
case, except at very early times, $\tau_{\star} \lesssim 1-3 \times
10^7$ yr.

@@

Figure~\ref{fig:vwSources} shows $\vwO(M_{\bullet})$ from each heating
source (stars, MSPs, SNe Ia, and black hole feedback) calculated for
the average star formation history of galaxies containing a given
black hole mass.  Again, Ia heating is suppressed in low mass galaxies
because the Ia radius exceeds the nominal stagnation radius (bottom
panel).  MSP heating is again found to be negligible for all black
hole masses.  The younger stellar populations characterizing galaxies
with $\Mbh\lsim 5\times 10^{8} \Msun$ are dominated by stellar wind
and Ia heating, with $v_{w}^{\star} \sim 1000$ km s$^{-1}$ and
$v_w^{\rm Ia} \sim 500$ km s$^{-1}$, respectively.  Only for the most
massive galaxies, with $\Mbh\gsim 5\times 10^{8} \Msun$, does the lack
of young stellar populations significantly reduce the role of stellar
wind
heating.  %Solid and dashed lines again show black hole masses for which the flow is thermally stable and thermally unstable, respectively, with $(\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|)_{r_{s}}$ as a function of $M_{\bullet}$ shown in the bottom panel.
Strikingly, {\it we find that galactic nuclei experiencing the same
  average star formation history as their host galaxies produce
  thermally stable flows across almost all black hole masses}.  One exception may occur for the highest mass black holes, for which the total heating velocity is comparable to the stellar velocity dispersion and thermal stability for the average star formation history is marginal ($\S\ref{sec:mdot}$).
{\bf AG: I wouldn't put too much stock in this since for higher masses
vw is comparable to sigma, and thus rs should run away to large
radii. This is especially true for core galaxies.}

Realistic variations (``burstiness") in the star formation history
will often result in lower heating rates.  Figure \ref{fig:NickPlot2}
shows the heating rate as a function of average black hole mass for
non-fiducial cases in which the current ($z = 0$) star formation is
suppressed by a factor $\zeta$ from its average $z = 0$ value, for a
characteristic timescale $t_{\rm scale}$ (for details, see Appendix
\ref{app:windheat}).  For $t_{\rm scale} \lesssim 10^{7}$ yr, we see
that $v_{\rm w}^{\star}$ is reduced by at most a factor of $\approx 2$
from its average value, even for huge drops in the star formation rate
($\zeta \sim 10^{-3}$).  However, burstiness in the star formation
history over longer timescales can suppress heating more
significantly.  When $t_{\rm scale} \gtrsim 10^8~{\rm yr}$, $v_{\rm
  w}^\star$ becomes very sensitive to $\zeta$: at $\zeta=0.1$, $v_{\rm
  w}^\star \approx 500~{\rm km~s}^{-1}$ in most galaxies, but for
smaller $\zeta$, $v_{\rm w}^\star \approx 200~{\rm km~s}^{-1}$ (an
exception to this is in galactic nuclei with $M_\bullet \gtrsim
5\times 10^8~M_\odot$, where heating is stabilized by SN Ia).  Our use
of average cosmic star formation histories is appropriate provided
local variations are either on short timescales ($t_{\rm scale}
\lesssim 10^7~{\rm yr}$), or limited in magnitude ($\zeta \gtrsim
0.1$).  If both of these conditions are violated, then the effective
heating rates for all but the most massive galaxies (where Ia
explosions dominate) will fall by a factor $\approx 5$ from our
fiducial volumetric averages, taken from \citet{MosterNaab+:2013a}.

>>>>

For low-mass black holes, stellar winds are the most important source of heating on
timescales of less than a few tens of millions of years.  Compton
heating comes to dominate at later times due to the higher accretion
rates that accompany the overall decrease in the heating rate, coupled
with the persistently high mass loss rates and gas densities associated with a relatively young
stellar population.  Compton heating dominates at all epochs for
higher mass black holes, with stellar winds and Ia SN competing for
the most significant non-AGN heating source.  As the total heating
rate declines with time, after some point the flow becomes thermally
unstable according to the criterion $(\dot{q}_{\rm heat}/|\dot{q}_{\rm
  rad}|)_{r_{s}} \lesssim 10$ (eq.~[\ref{eq:cooling3}]) {\bf BDM: correct?}, as shown by
dashed lines in Fig.~\ref{fig:vwSourcesImp}.  The thermal stability threshold is
calculated neglecting Compton heating because SMBH feedback may not be
capable of truly stabilizing the flow given its inability to respond
instantaneously to local changes in the properties of the gas.
Thermally unstable flow is seen to be the norm in the single starburst
case, except at very early times, $\tau_{\star} \lesssim 1-3 \times
10^7$ yr.

@@

Figure~\ref{fig:vwSources} shows $\vwO(M_{\bullet})$ from each heating
source (stars, MSPs, SNe Ia, and black hole feedback) calculated for
the average star formation history of galaxies containing a given
black hole mass.  Again, Ia heating is suppressed in low mass galaxies
because the Ia radius exceeds the nominal stagnation radius (bottom
panel).  MSP heating is again found to be negligible for all black
hole masses.  The younger stellar populations characterizing galaxies
with $\Mbh\lsim 5\times 10^{8} \Msun$ are dominated by stellar wind
and Ia heating, with $v_{w}^{\star} \sim 1000$ km s$^{-1}$ and
$v_w^{\rm Ia} \sim 500$ km s$^{-1}$, respectively.  Only for the most
massive galaxies, with $\Mbh\gsim 5\times 10^{8} \Msun$, does the lack
of young stellar populations significantly reduce the role of stellar
wind
heating.  %Solid and dashed lines again show black hole masses for which the flow is thermally stable and thermally unstable, respectively, with $(\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|)_{r_{s}}$ as a function of $M_{\bullet}$ shown in the bottom panel.
Strikingly, {\it we find that galactic nuclei experiencing the same
  average star formation history as their host galaxies produce
  thermally stable flows across almost all black hole masses}.  One exception may occur for the highest mass black holes, for which the total heating velocity is comparable to the stellar velocity dispersion ($\zeta \sim \zeta_c$) and thermal stability for the average star formation history is marginal, especially for core galaxies ($\S\ref{sec:mdot}$).

Realistic variations (``burstiness") in the star formation history
will often result in lower heating rates.  Figure \ref{fig:NickPlot2}
shows the heating rate as a function of average black hole mass for
non-fiducial cases in which the current ($z = 0$) star formation is
suppressed by a factor $\iota$ from its average $z = 0$ value, for a
characteristic timescale $t_{\rm scale}$ (for details, see Appendix
\ref{app:windheat}).  For $t_{\rm scale} \lesssim 10^{7}$ yr, we see
that $v_{\rm w}^{\star}$ is reduced by at most a factor of $\approx 2$
from its average value, even for huge drops in the star formation rate
($\iota \sim 10^{-3}$).  However, burstiness in the star formation
history over longer timescales can suppress heating more
significantly.  When $t_{\rm scale} \gtrsim 10^8~{\rm yr}$, $v_{\rm
  w}^\star$ becomes very sensitive to $\iota$: at $\iota=0.1$, $v_{\rm
  w}^\star \approx 500~{\rm km~s}^{-1}$ in most galaxies, but for
smaller $\iota$, $v_{\rm w}^\star \approx 200~{\rm km~s}^{-1}$ (an
exception to this is in galactic nuclei with $M_\bullet \gtrsim
5\times 10^8~M_\odot$, where heating is stabilized by SN Ia).  Our use
of average cosmic star formation histories is appropriate provided
local variations are either on short timescales ($t_{\rm scale}
\lesssim 10^7~{\rm yr}$), or limited in magnitude ($\iota \gtrsim
0.1$).  If both of these conditions are violated, then the effective
heating rates for all but the most massive galaxies (where Ia
explosions dominate) will fall by a factor $\approx 5$ from our
fiducial volumetric averages, taken from \citet{MosterNaab+:2013a}.

### 2015-03-31 windheat.tex
blah blah

 \begin{figure}
  \includegraphics[width=\columnwidth]{NickPlot2.pdf}
  \caption{\label{fig:NickPlot2} Effective wind velocities for nonstandard
    star formation histories.  The black curve shows, for reference,
    $v_{\rm w}$ calculated using the halo-averaged $S(t)$.  Other
    colors show the effective wind velocities $\mathcal{V}_{\rm w}$
    computed using convolved star formation histories
    $\mathcal{S}(t)$.  The purple curve shows a recent drop in the
    star formation rate, i.e. $t_{\rm scale}=10^6~{\rm yr}$.  The blue
    and red curves show $t_{\rm scale}=10^7~{\rm yr}$ and $t_{\rm
      scale}=10^8~{\rm yr}$, respectively; longer scale times look
    quite similar to the red curves.  Dashed, dotted, and dot-dashed
    line styles show star formation rates that have declined to
    fractions $\zeta = 0.001$, $\zeta = 0.1$, and $\zeta = 0.3$ of
    their halo-averaged values, respectively. }
  \end{figure}
  We estimate the stellar wind heating provided by the {\it average}
  star formation history of galaxies of a given $\Mbh$ using the
  results of \citet[eqs.~17-20]{MosterNaab+:2013a}.  Note that the
  star formation histories in \citet{MosterNaab+:2013a} are in terms
  of halo mass. For a given $\Mbh$ we assign the halo mass whose star
  formation history would produce a bulge consistent with the
  $\Mbh-M_{\rm bulge}$ relationship from \citet{McConnellMa:2013a}.  A
  slight complication occurs for the largest mass halos, where much of
  the $z=0$ stellar mass has been acquired through accretion of
  satellite halos rather than {\it in situ} star formation.  To
  accomodate this, we incorporate analytic fits for mass accretion
  histories, taken from \citet[their eqs.~21-23]{MosterNaab+:2013a},
  assuming that the age distribution of the accreted stars is equal to
  the age distribution of those formed {\it in situ}.  This assumption
  may be conservative if the primary galaxy's accretion history is
  dominated by minor mergers with younger stellar populations.  On the
  other hand, the dynamical friction inspiral time for small satellite
  galaxies is quite long, generally much greater than the $\sim
  10^7~{\rm yr}$ for which young stars can dominate the heating
  budget.

@@

Figure \ref{fig:NickPlot2} shows how the wind heating varies as star formation histories deviate from their halo-averaged values.  In particular, we show
\begin{align} 
  \dot{\mathcal{M}}(t) &= \int_0^t \mathcal{S}(t_1) \dot{\bar{m}}(t-t_1){\rm
      d}t_1\\
  \dot{\mathcal{E}}(t) &= \int_0^t \mathcal{S}(t_1) \dot{\bar{e}}(t-t_1){\rm
      d}t_1,\\
  \mathcal{V}_w^2(t) &=2 \dot{\mathcal{E}}(t)/\dot{\mathcal{M}}(t).
\end{align}
In these equations, 
\begin{equation}
\mathcal{S}(t) = S(t) \times \left(\frac{2}{\pi}(1-\zeta)\arctan(t/t_{\rm scale}) + \zeta \right).
\end{equation}
This function convolves the recent ($z \approx 0$) halo-averaged star formation history with local variation to give a more pessimistic estimate for the value of $\tilde{v}_{\rm w}$.  In particular, replacing $S(t)$ with $\mathcal{S}(t)$ reduces the recent star formation rate to a fraction $\epsilon$ of its halo-averaged value, and does so for a characteristic time $t_{\rm scale}$ into the past.  As we can see in Fig. \ref{fig:NickPlot2}, this dramatically lowers the effective wind speed when both $t_{\rm scale} \gtrsim 10^7 ~{\rm yr}$ and $\epsilon \lesssim 0.1$, but otherwise has too modest of an effect to change the thermal stability properties of the flow (although the location of $r_{\rm s}$ and the value of $\dot{M}$ may change significantly).

>>>>

 \begin{figure}
  \includegraphics[width=\columnwidth]{NickPlot2.pdf}
  \caption{\label{fig:NickPlot2} Effective wind velocities for nonstandard
    star formation histories.  The black curve shows, for reference,
    $v_{\rm w}$ calculated using the halo-averaged $S(t)$.  Other
    colors show the effective wind velocities $\mathcal{V}_{\rm w}$
    computed using convolved star formation histories
    $\mathcal{S}(t)$.  The purple curve shows a recent drop in the
    star formation rate, i.e. $t_{\rm scale}=10^6~{\rm yr}$.  The blue
    and red curves show $t_{\rm scale}=10^7~{\rm yr}$ and $t_{\rm
      scale}=10^8~{\rm yr}$, respectively; longer scale times look
    quite similar to the red curves.  Dashed, dotted, and dot-dashed
    line styles show star formation rates that have declined to
    fractions $\iota= 0.001$, $\iota = 0.1$, and $\iota = 0.3$ of
    their halo-averaged values, respectively. }
  \end{figure}
  We estimate the stellar wind heating provided by the {\it average}
  star formation history of galaxies of a given $\Mbh$ using the
  results of \citet[eqs.~17-20]{MosterNaab+:2013a}.  Note that the
  star formation histories in \citet{MosterNaab+:2013a} are in terms
  of halo mass. For a given $\Mbh$ we assign the halo mass whose star
  formation history would produce a bulge consistent with the
  $\Mbh-M_{\rm bulge}$ relationship from \citet{McConnellMa:2013a}.  A
  slight complication occurs for the largest mass halos, where much of
  the $z=0$ stellar mass has been acquired through accretion of
  satellite halos rather than {\it in situ} star formation.  To
  accomodate this, we incorporate analytic fits for mass accretion
  histories, taken from \citet[their eqs.~21-23]{MosterNaab+:2013a},
  assuming that the age distribution of the accreted stars is equal to
  the age distribution of those formed {\it in situ}.  This assumption
  may be conservative if the primary galaxy's accretion history is
  dominated by minor mergers with younger stellar populations.  On the
  other hand, the dynamical friction inspiral time for small satellite
  galaxies is quite long, generally much greater than the $\sim
  10^7~{\rm yr}$ for which young stars can dominate the heating
  budget.

@@

Figure \ref{fig:NickPlot2} shows how the wind heating varies as star formation histories deviate from their halo-averaged values.  In particular, we show
\begin{align} 
  \dot{\mathcal{M}}(t) &= \int_0^t \mathcal{S}(t_1) \dot{\bar{m}}(t-t_1){\rm
      d}t_1\\
  \dot{\mathcal{E}}(t) &= \int_0^t \mathcal{S}(t_1) \dot{\bar{e}}(t-t_1){\rm
      d}t_1,\\
  \mathcal{V}_w^2(t) &=2 \dot{\mathcal{E}}(t)/\dot{\mathcal{M}}(t).
\end{align}
In these equations, 
\begin{equation}
\mathcal{S}(t) = S(t) \times \left(\frac{2}{\pi}(1-\iota)\arctan(t/t_{\rm scale}) + \iota \right).
\end{equation}
This function convolves the recent ($z \approx 0$) halo-averaged star formation history with local variation to give a more pessimistic estimate for the value of $\tilde{v}_{\rm w}$.  In particular, replacing $S(t)$ with $\mathcal{S}(t)$ reduces the recent star formation rate to a fraction $\epsilon$ of its halo-averaged value, and does so for a characteristic time $t_{\rm scale}$ into the past.  As we can see in Fig. \ref{fig:NickPlot2}, this dramatically lowers the effective wind speed when both $t_{\rm scale} \gtrsim 10^7 ~{\rm yr}$ and $\epsilon \lesssim 0.1$, but otherwise has too modest of an effect to change the thermal stability properties of the flow (although the location of $r_{\rm s}$ and the value of $\dot{M}$ may change significantly).
